Idealized Wind Models of Cool Gas Outflows

### 2010-09-30 Winds/papers/Theory/First/Submit/ms.tex
profx
git-svn-id: https://svn.ucolick.org/xsvn/trunk/radtrans@5251 ac11df5e-a788-c23b-a8e1-c758c323924d

\author{
J. Xavier Prochaska\altaffilmark{1}, 
Daniel Kasen\altaffilmark{1,2,3}, 
Kate Rubin\altaffilmark{1} 
}
\altaffiltext{1}{Department of Astronomy and Astrophysics, UCO/Lick Observatory, University of California, 1156 High Street, Santa Cruz, CA 95064}
\altaffiltext{2}{UCB?}
\altaffiltext{3}{Hubble Fellow}

@@

Iron exhibits the most complex set of energy levels for elements
frequently studied in astrophysics.  The Fe$^+$ ion alone has over
10,000 energy levels recorded \citep{iron}, and even this is an
incomplete list.  
One reason for iron's complexity is
that the majority of its configurations exhibit fine-structure splitting.
This includes the ground-state configuration (\aconfig) which is split
into 5 levels, 
labeled by the total angular momentum $J$, 
with excitation energies $T_{\rm ex} \equiv \Delta E / k$ ranging from
$T_{\rm ex} \approx 500-1500$\,K (Figure~\ref{fig:energy}).  
Transitions between these fine-structure levels are 
forbidden (magnetic-dipole) and have spontaneous decay times of several hours.

@@

For a wind with a steep velocity law (i.e.\ a large gradient $dv/dr$)
and/or a narrow intrinsic profile (i.e.\ a small Doppler parameter),
\citet{sobolev60} introduced a formalism that defines the opacity of
the wind at a given radius $\tau(r)$ in terms of the density and
velocity gradient of the flow at that radius;

\begin{equation}
\tau^S(r) = \kappa^\ell \rho(r) \, \lambda^\ell \ltp \frac{dv}{dr} 
\rtp^{-1}  \cmma
\label{eqn:Sobolev}
\end{equation}
where $\kappa^\ell$ is the opacity of transition $\ell$.  This opacity
applies to a photon with wavelength $\lambda^S
\equiv \lambda^\ell (1-v(r)/c)$.
This is, of course, not the true opacity;
a photon with wavelength $\lambda^S$ that travels through the wind 
experiences a total opacity 
$\tau_\nu^\ell = \intl_0^\infty \rho(r) \kappa_\nu^\ell dr$.  The Sobolev
approximation is valid when the opacity the photon experiences is
dominated by gas in a localized portion of the flow.
We find that it applies to nearly all of the
models presented in this paper and therefore provides a convenient
approach to estimating $\tau^\ell_\nu$.

@@

\begin{equation}
n_{\rm H} (r) = \frac{n^0_{\rm H}}{r^2} \;\;\; , 
\label{eqn:density}
\end{equation}
and a velocity law with a purely radial flow

\begin{equation}
\vec v = v_r (r) \hat r = v_0 r \hat r  \perd
\label{eqn:vel}
\end{equation}
Turbulent motions are
characterized by a Doppler parameter\footnote{
  The adopted Doppler parameter
  ($\mbturb = 15 \mkms$) has a minor impact on the results.
  The absorption profiles are insensitive to its value and varying \bturb\
  only tends to modify the widths and modestly
  shift the centroids of emission lines.} 
, $b_{\rm turb}$.  
We convert the hydrogen density $n_{\rm H}$ to the number densities of
Mg$^+$ and Fe$^+$ ions by assuming solar relative abundances with an
absolute metallicity of 1/2 solar and depletion factors of 1/10 and
1/20 for Mg and Fe respectively, i.e.\  $\mnmg = 10^{-5.47} n_{\rm H}$ 
and \nfe=\nmg/2. At the center of the wind is a homogeneous source of
continuum photons with size $r_{\rm source}$. The parameters for the
fiducial wind model are summarized in Table~\ref{tab:fiducial}.

@@

Because the results are dictated by frequency dependence of the
$\tau_{2796}$ profile, 
the actual dimensions and density of the wind are
essentially unimportant provided they scale together to give nearly the same
optical depth. Therefore, one may consider the choices for
$r_{\rm inner}, r_{\rm outer}, n_{\rm H}^0$ as arbitrary.
Nevertheless, we adopted values for this fiducial model with
some astrophysical motivation,  e.g., values that correspond to
galactic dimensions and a normalization that gives $\tau_{2796}^{\rm
  max} \sim 10$.

@@

The fiducial model assumes an
isotropic wind with only radial variations in velocity and density. 
Angular isotropy is obviously an idealized case, but
it is frequently assumed in studies of galactic-scale outflows
\citep[e.g.][]{steidel+10}.   There are several reasons, however, to
consider anisotropic winds.  Firstly, galaxies are not spherically
symmetric;  the sources driving the
wind (e.g.\ supernovae, AGN) are very unlikely to be isotropically distributed
within the galaxy.  
Secondly, the galactic ISM frequently has a disk-like morphology
which will suppress the wind preferentially at low galactic latitudes,
perhaps yielding a bi-conic morphology \citep[e.g.][]{ham90,wws02}.
Lastly, the galaxy could be surrounded by an
anisotropic gaseous halo whose interaction would produce an irregular 
outflow.

@@

The differences in the \ion{Fe}{2} absorption profiles are reflected
in the much higher strengths ($3-10\times$) of emission from
transitions to the excited states of \aconfig.   This occurs because:
(1) there is greater absorption by the \feiid\
resonance lines; (2) the high opacity of the ISM component leads to
an enhanced conversion of resonance photons with $\mdv \approx 0\mkms$
into \feiis\ photons.  This is especially notable for the
\feiib\ transition whose coupled \feiis\ transition shows an equivalent width nearly
$10\times$ stronger than for the fiducial model.  The relative
strengths of the \feiib\ and \feiie\ lines provide a direct
diagnostic on the degree to which the resonance lines are trapped.
[Comment:  this is a product of $\tau_{2600}$ and the velocity
gradient]

@@

In the following, we consider two methods to analyze the LBG wind.
Both approaches assume an isotropic wind and adopt the velocity law given by
Equation~\ref{eqn:LBG_vlaw}.  In one model, we treat the cool gas as a
diffuse medium with unit covering fraction and a radial density
profile determined from the Sobolev approximation.  We then apply 
the Monte Carlo methodology used for the other wind models to predict
\ion{Mg}{2} and \ion{Fe}{2} line profiles.  In the other LBG model,
we modify our algorithms to more precisely mimic the concept of an
ensemble of optically thick clouds with a partial covering fraction on
galactic scales.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Sobolev approximation}
\label{sec:Sobolev}

As demonstrated in Figure~\ref{fig:LBG_Sobolev}, the wind velocity for
the LBG model rises very steeply with increasing radius before
flattening at large radii.  Under these conditions, the Sobolev
approximation provides an accurate description of the optical depth
for the flowing medium. 
In our case, the optical depth profile is actually provided by 
Equation~\ref{eqn:LBG_I},

\begin{equation}
\tau_{\rm LBG}(v) = -\ln \ltk I_{\rm LBG}(v) \rtk \perd
\label{eqn:tauLBG}
\end{equation}
Therefore, we may invert the Sobolev approximation (Equation~\ref{eqn:Sobolev})
to derive the radial density profile,

\begin{equation}
n_{\rm LBG}(r) = \frac{\tau_{\rm LBG}(r) \; d\mvr/dr}{\kappa^\ell
  \lambda^\ell} \cmma
\label{eqn:nSobolev}
\end{equation}
with $\lambda^\ell$ the rest wavelength and $\kappa^\ell \equiv f^\ell
\pi^2 e^2/m_e c$.

@@

In the previous sub-section, we described a Sobolev solution that
reproduces the average absorption profile of LBGs in cool gas
transitions when scattered photons are ignored.  A proper analysis
that includes scattered photons,
however, predicts line-profiles that are qualitatively
different from the observations because scattered photons fill-in
absorption and generate significant line-emission (similar to the
fiducial wind model; $\S$~\ref{sec:fiducial}).
We also argued that the implied mass, energy, and momentum
distributions of this wind (Figure~\ref{fig:LBG_cumul} were too
extreme. 
This LBG-Sobolev model, however, is not precisely
the one proposed by S10;  those authors proposed an ensemble of optically
thick clouds with a partial covering fraction described by
Equation~\ref{eqn:covering}.  In contrast, the LBG-Sobolev model assumes
a diffuse medium with a declining density profile but a unit covering
fraction.  One may question, therefore,  whether the differences
lead to the failures of the LBG-Sobolev model.

>>>>

\author{
J. Xavier Prochaska\altaffilmark{1}, 
Daniel Kasen\altaffilmark{2,3}, 
Kate Rubin\altaffilmark{1} 
}
\altaffiltext{1}{Department of Astronomy and Astrophysics, UCO/Lick Observatory, University of California, 1156 High Street, Santa Cruz, CA 95064}
\altaffiltext{2}{Departments of Physics and Astronomy, University of California, Berkeley, 366 LeConte Hall, Berkeley CA, 94720}
\altaffiltext{3}{Nuclear Science Division, Lawrence Berkeley National Laboratory, 1 Cyclotron Rd, Berkeley, CA, 94720}

@@

Iron exhibits the most complex set of energy levels for elements
frequently studied in astrophysics.  The Fe$^+$ ion alone has 
millions of energy levels recorded \citep{kurucz_cd1}, and even this is an
incomplete list.  
One reason for iron's complexity is
that the majority of its configurations exhibit fine-structure splitting.
This includes the ground-state configuration (\aconfig) which is split
into 5 levels, 
labeled by the total angular momentum $J$, 
with excitation energies $T_{\rm ex} \equiv \Delta E / k$ ranging from
$T_{\rm ex} \approx 500-1500$\,K (Figure~\ref{fig:energy}).  
Transitions between these fine-structure levels are 
forbidden (magnetic-dipole) and have spontaneous decay times of several hours.

@@

A photon propagating through a differentially expanding medium
interacts with a line only when
it's comoving frame wavelength is Doppler shifted into near resonance with the line center rest wavelength. 
For a wind with a steep velocity law  (i.e.\ a large gradient $dv/dr$)
and/or a narrow intrinsic profile (i.e.\ a small Doppler parameter), the 
spatial extent of the region of resonance
may by much smaller than the length scale of the
wind itself.  The interaction can then 
be considered to occur at  a point.
In this case, \citet{sobolev60} introduced a formalism that gives the line optical depth $\tau(r)$ at  
a given point in terms of the density and
velocity gradient of the flow at that radius.  For a wind in homologous expansion,
this optical depth is independent of the direction of propagation and given by
\begin{equation}
\tau^S(r) = K^\ell   \lambda_0^\ell | dv/dr |^{-1} \cmma
\label{eqn:Sobolev}
\end{equation}
where 
\begin{equation}
K^\ell = \frac{\pi e^2}{m_e c} f^\ell n^\ell(r)
\end{equation}
is the integrated line opacity, $\lambda_0^\ell$ the rest center wavelength,
$f^\ell$ the oscillator strength and
$n^\ell$ the density in the lower level of the transition.   We have
neglected corrections for stimulated emission.   This optical
depth applies to a photon with wavelength $\lambda^S
\equiv \lambda^\ell (1- v(r)~\hat{r}  \cdot \hat{d} /c)$ where $\hat{d}$ is the 
direction of propagation.  The probability that such a photon
is scattered/absorbed at the point of resonance is simply $1 - \exp[-\tau^S(r)]$.
We find that the Sobolev approximation applies for nearly all of the
models presented in this paper, and therefore provides a convenient
approach to estimating the optical depth.

@@

\begin{equation}
n_{\rm H} (r) = n^0_{\rm H} \biggl( \frac{r_{\rm inner}}{r}  \biggr)^2\;\;\; , 
\label{eqn:density}
\end{equation}
and a velocity law with a purely radial flow

\begin{equation}
\vec v = v_r (r) \hat r = \frac{v_0 r}{r_{\rm outer}} \hat r  \perd
\label{eqn:vel}
\end{equation}
For the fiducial case,  the hydrogen density at
the inner radius is $n_{\rm H}^0 = 0.1~{\rm cm}^{-3}$ and the
velocity at the outer radius is $v_0 = 1,000$~\kms. Turbulent motions are
characterized by a Doppler parameter\footnote{
  The adopted Doppler parameter
 has a minor impact on the results.
  The absorption profiles are insensitive to its value and varying \bturb\
  only tends to modify the widths and modestly
  shift the centroids of emission lines.} 
$b_{\rm turb} = 15 \mkms$.  
We convert the hydrogen density $n_{\rm H}$ to the number densities of
Mg$^+$ and Fe$^+$ ions by assuming solar relative abundances with an
absolute metallicity of 1/2 solar and depletion factors of 1/10 and
1/20 for Mg and Fe respectively, i.e.\  $\mnmg = 10^{-5.47} n_{\rm H}$ 
and \nfe=\nmg/2. At the center of the wind is a homogeneous source of
continuum photons with size $r_{\rm source}$. The parameters for the
fiducial wind model are summarized in Table~\ref{tab:fiducial}.

@@

%Because the results are dictated by frequency dependence of the
% $\tau_{2796}$ profile
%  I don't understand this sentence so I commented it out -- ? -- Dan 
For the spatial integrated (i.e., 1D) spectra, the actual dimensions and density of the wind are
unimportant provided they scale together to give nearly the same
optical depth. Therefore, one may consider the choices for
$r_{\rm inner}, r_{\rm outer}, n_{\rm H}^0$ as arbitrary.
Nevertheless, we adopted values for this fiducial model with
some astrophysical motivation,  e.g., values that correspond to
galactic dimensions and a normalization that gives $\tau_{2796}^{\rm
  max} \sim 10$.

@@

The fiducial model assumes an
isotropic wind with only radial variations in velocity and density. 
Angular isotropy is obviously an idealized case, but
it is frequently assumed in studies of galactic-scale outflows
\citep[e.g.][]{steidel+10}.   There are several reasons, however, to
consider anisotropic winds.  Firstly, galaxies are not spherically
symmetric;  the sources driving the
wind (e.g.\ supernovae, AGN) are very unlikely to be spherically distributed
within the galaxy.  
Secondly, the galactic ISM frequently has a disk-like morphology
which will suppress the wind preferentially at low galactic latitudes,
perhaps yielding a bi-conic morphology \citep[e.g.][]{ham90,wws02}.
Lastly, the galaxy could be surrounded by an
aspherical gaseous halo whose interaction would produce an irregular 
outflow.

@@

The differences in the \ion{Fe}{2} absorption profiles are reflected
in the much higher strengths ($3-10\times$) of emission from
transitions to the excited states of \aconfig.   This occurs because:
(1) there is greater absorption by the \feiid\
resonance lines; (2) the high opacity of the ISM component leads to
an enhanced conversion of resonance photons with $\mdv \approx 0\mkms$
into \feiis\ photons.  This is especially notable for the
\feiib\ transition whose coupled \feiis\ transition shows an equivalent width nearly
$10\times$ stronger than for the fiducial model.  The relative
strengths of the \feiib\ and \feiie\ lines provide a direct
diagnostic on the degree to which the resonance line photons are trapped.
[Comment:  this is a product of $\tau_{2600}$ and the velocity
gradient]

@@

In the following, we consider two methods to analyze the LBG wind.
Both approaches assume an isotropic wind and adopt the velocity law given by
Equation~\ref{eqn:LBG_vlaw}.  In the first model, we treat the cool gas as a
diffuse medium with unit covering fraction and a radial density
profile determined from the Sobolev approximation.  We then apply 
the Monte Carlo methodology used for the other wind models to predict
\ion{Mg}{2} and \ion{Fe}{2} line profiles.  In the second LBG model,
we modify our algorithms to more precisely mimic the concept of an
ensemble of optically thick clouds with a partial covering fraction on
galactic scales.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Sobolev Inversion}
\label{sec:Sobolev}

As demonstrated in Figure~\ref{fig:LBG_Sobolev}, the wind velocity for
the LBG model rises very steeply with increasing radius before
flattening at large radii.  Under these conditions, the Sobolev
approximation provides an accurate description of the optical depth
for the flowing medium. 
The Sobolev line optical depth profile can be derived from
the absorption profile (Equation~\ref{eqn:LBG_I}), 
\begin{equation}
\tau_{\rm LBG}(v) = -\ln \ltk I_{\rm LBG}(v) \rtk \perd
\label{eqn:tauLBG}
\end{equation}
This simple expression assumes that the source size is small compared the wind dimensions
and neglects the effects of light scattered in the wind (for more general inversion
formulae that do not make these assumptions, see \citet{Kasen_2002}).
The Sobolev approximation (Equation~\ref{eqn:Sobolev}) provides a simple expression for the optical
depth at each radius in terms of the local density and velocity gradient.  In general, the velocity 
 gradient must be taken along the direction of propagation of a photon, however
in this model the central source is of relatively small  and the radiation
field is therefore primarily radially directed.  In this 
case the velocity gradient in the radial direction, $d\mvr/dr$, is the relevant quantity, and 
we may use Equation~\ref{eqn:Sobolev}  to invert
the optical depth and determine the density profile,
\begin{equation}
n_{\rm LBG}(r) =  \tau_{\rm LBG}(r)  \frac{ |d\mvr/dr |}{f^\ell  \lambda_0^\ell}
\biggl( \frac{ m_e c}{\pi e^2} \biggl)
 \cmma
\label{eqn:nSobolev}
\end{equation}

@@

In the previous sub-section, we described a Sobolev inversion that
reproduces the average absorption profile of LBGs in cool gas
transitions when scattered photons are ignored.  A proper analysis
that includes scattered photons,
however, predicts line-profiles that are qualitatively
different from the observations because scattered photons fill-in
absorption and generate significant line-emission (similar to the
fiducial wind model; $\S$~\ref{sec:fiducial}).
We also argued that the implied mass, energy, and momentum
distributions of this wind (Figure~\ref{fig:LBG_cumul}) were too
extreme. 
This LBG-Sobolev model, however, is not precisely
the one proposed by S10;  those authors proposed an ensemble of optically
thick clouds with a partial covering fraction described by
Equation~\ref{eqn:covering}.  In contrast, the LBG-Sobolev model assumes
a diffuse medium with a declining density profile but a unit covering
fraction.  One may question, therefore,  whether the differences
lead to the failures of the LBG-Sobolev model.
